Regression Cube Analysis of Cohort Study Data

### 2015-03-19 draft.tex
Added workflow subsec to Cube section

%% Abstract section.
\abstract{%
\com{Problem}.
Epidemiological studies comprise heterogeneous data about a subject group (a \emph{cohort}) to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subjects' lifestyle, medical conditions and also medical image data.
%%
These features are analyzed using statistical regression analysis to identify feature combinations indicating a disease (the \emph{target feature}).
%%
Although there is a strong demand for overview visualizations of a whole data set towards a target feature, no suitable tool is available for epidemiological researchers.

\com{New Solution}.
%%
We propose an analysis approach of epidemiological data sets by incorporating all features in an exhaustive regression-based analysis.
%%
This approach combines all \emph{independent features} with respect to a \emph{target feature} and provides a visualization, which reveals insight into the data by highlighting relationships.
%%
A 3D-visualization of all combinations of two to three independent features towards a target acts as an overview of the whole data set, the \emph{Regression Cube}.
%%
Slicing through the \emph{Regression Cube} allows for the detailed analysis of features towards the target disease.
%%
Expert knowledge about disease-specific hypotheses can be included into the analysis by adjusting the regression model formulas.
%%
Furthermore, the influences of features can be assessed using a difference view comparing different calculation results.
%%
\com{Validation \& Results}.
%%
We applied our \emph{Regression Cube} method to a hepatic steatosis data set to reproduce results from a data-mining driven analysis.
%%
A qualitative analysis with three domain experts was conducted on a breast fat data set.
%%
We were able to derive new hypotheses about relations between breast fat density and breast lesions towards breast cancer.
%%
\com{Implications}.
%%
With our work, we present a visual overview of epidemiological data with the \emph{Regression Cube}, which allows for the first time a interactive regression-based analysis of large feature sets with respect to a disease.
} % end of abstract

@@

\subsection{Abstracting Regression Results using R$^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/cube_sketch}
 \caption{
 (a) Overview visualization using a heatmap for the formula $Z \sim X + Y$, where $Z$ assumes feature $Age$.
 %%
 The R$^2$ values extracted from the regression formulas depict the goodness of fit and are mapped to color saturation.
 %%
 A saturated color shows a strong correlation.
 %%
 (b) Since $Z$ assumes all features $n$ as given by the formula, it yields $n$ heatmap visualizations.
 %%
 These represent the slices in our cube visualization.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of a overview visualization is to provide a comprehensive view on the data, which is easy to understand.
%% TODO: Cite another paper here!
As described in \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded using color on a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the goodness of fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the R$^2$ value is the metric allowing for this kind of assessment.
%%
Since it is scaled between [0,1] it also allows for comparison between regression models.
%%
We can apply the same information mapping by translating the R$^2$ values to color saturation (Fig.~\ref{fig:Cube} (a)).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g. $Age \sim X + Y$).
%%
Introducing $Z$ creates a 3D \emph{Regression Cube}.
%%
The visualization of R$^2$ values derived from different regression cubes (e.g. $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the R$^2$ results of different regression methods are encoded using different colors (e.g. blue for linear regression and red for logistic regression).
%%
This way, the cube can easily be extended using other regression types.
%%
For cubes having one fixed target feature, such as $Cancer \sim X + Y + Z$ no such encodings is required and the $z$ dimension can be compared directly.

The goal is to create a overview visualization for a data set, but on the other hand we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
Different analysis approaches require different starting points using the \emph{Regression Cube}.

\subsection{Analysis Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_sketch}
 \caption{
 Workflow of the analysis using \emph{Regression Cubes}.
 %%
 The analysis starts with a set of features given by the data set.
 %%
 Then, the user may specify a formula regarding to specify a hypothesis, or use a predefined formula to start a explorative analysis of the data set.
 %%
 The Regression Cube is then visualized, where the user has either the option to adjust the formula or to derive details on demand on a specific regression.
 %%
 Insights into the data yield either an adjustment of the current formula.
 %%
 By applying a difference view, regression cubes can be compared to each other.
 %%
 The analysis yields insights and hypothesis about feature relations.
 %%
 }
  \label{fig:Workflow}
\end{figure}

\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
 \com{Workflow for different analysis approaches.}
 \com{0. Select the Variables.}
 \com{1. Select the Formulas}
 \com{a) Analysis without hypothesis about the data set (z ~ x+y)}
 \com{b) Analysis without hypothesis with confounder (z ~ age)}
 \com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
 \com{d) Use Difference View}
 \com{d) Clarify using confounder (Var ~ x)}
 \com{3. Find interesting pane.}
 \com{4. Get Details on Demand.}
%%
Analyze first, Show Overview, Details on Demand
%%
\section{System Design and Implementation} \label{Interaction- and Visualization Techniques}
%%

>>>>

%% Abstract section.
\abstract{%
%\com{Problem}.
Epidemiological studies comprise heterogeneous data about a subject group (a \emph{cohort}) to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subjects' lifestyle, medical conditions and also medical image data.
%%
These features are analyzed using statistical regression analysis to identify feature combinations indicating a disease (the \emph{target feature}).
%%
Although there is a strong demand for overview visualizations of a whole data set towards a target feature, no suitable tool is available for epidemiological researchers.

%\com{New Solution}.
%%
We propose an analysis approach of epidemiological data sets by incorporating all features in an exhaustive regression-based analysis.
%%
This approach combines all \emph{independent features} with respect to a \emph{target feature} and provides a visualization, which reveals insight into the data by highlighting relationships.
%%
A 3D-visualization of all combinations of two to three independent features towards a target acts as an overview of the whole data set, the \emph{Regression Cube}.
%%
Slicing through the \emph{Regression Cube} allows for the detailed analysis of features towards the target disease.
%%
Expert knowledge about disease-specific hypotheses can be included into the analysis by adjusting the regression model formulas.
%%
Furthermore, the influences of features can be assessed using a difference view comparing different calculation results.
%%
%\com{Validation \& Results}.
%%
We applied our \emph{Regression Cube} method to a hepatic steatosis data set to reproduce results from a data-mining driven analysis.
%%
A qualitative analysis with three domain experts was conducted on a breast fat data set.
%%
We were able to derive new hypotheses about relations between breast fat density and breast lesions towards breast cancer.
%%
%\com{Implications}.
%%
With our work, we present a visual overview of epidemiological data with the \emph{Regression Cube}, which allows for the first time a interactive regression-based analysis of large feature sets with respect to a disease.
} % end of abstract

@@

\subsection{Abstracting Regression Results using R$^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/cube_sketch}
 \caption{
 (a) Overview visualization using a heatmap for the formula $Z \sim X + Y$, where $Z$ assumes feature $Age$.
 %%
 The R$^2$ values extracted from the regression formulas depict the goodness of fit and are mapped to color saturation.
 %%
 A saturated color shows a strong correlation.
 %%
 (b) Since $Z$ assumes all features $n$ as given by the formula, it yields $n$ heatmap visualizations.
 %%
 These represent the slices in our cube visualization.
 %%
 The R$^2$ values of each slice voxel is mapped on opacity in the 3D-view, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of a overview visualization is to provide a comprehensive view on the data, which is easy to understand.
%% TODO: Cite another paper here!
As described in \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded using color on a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the goodness of fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the R$^2$ value is the metric allowing for this kind of assessment.
%%
\paragraph{2D (Slice) View.}
%%
Since R$^2$ is scaled between [0,1] it also allows for comparison between regression models.
%%
We can apply the same mosaic plot mapping by translating the R$^2$ values to color saturation (Fig.~\ref{fig:Cube} a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g. $Age \sim X + Y$).
%%

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
R$^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of R$^2$ values derived from different regression cubes (e.g. $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the R$^2$ results of different regression methods are encoded using different colors (e.g. blue for linear regression and red for logistic regression).
%%
This way, the cube can easily be extended using other regression types.
%%
For cubes having one fixed target feature, such as $Cancer \sim X + Y + Z$ no such encodings is required and the $z$ dimension can be compared directly.
%%
\\\\
The goal is to create a overview visualization for a data set, but on the other hand we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
Different analysis approaches require different starting points using the \emph{Regression Cube}.

\subsection{Analysis Workflow}
\com{I'm not satisfied with the structure in this subsection, please proofread it with focus in the train of thought!}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_sketch}
 \caption{
 Workflow of the analysis using \emph{Regression Cubes}.
 %%
 The analysis starts with a set of features given by the data set.
 %%
 Then, the user may specify a formula regarding to specify a hypothesis, or use a predefined formula to start a explorative analysis of the data set.
 %%
 The Regression Cube is then visualized, where the user has either the option to adjust the formula or to derive details on demand on a specific regression.
 %%
 Insights into the data yield either an adjustment of the current formula.
 %%
 By applying a difference view, regression cubes can be compared to each other.
 %%
 The analysis yields insights and hypothesis about feature relations.
 %%
 }
  \label{fig:Workflow}
\end{figure}
%%
%\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%\com{Workflow for different analysis approaches.}
%\com{0. Select the Variables.}
%\com{1. Select the Formulas}
%\com{a) Analysis without hypothesis about the data set (z ~ x+y)}
%\com{b) Analysis without hypothesis with confounder (z ~ age)}
%\com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
%\com{d) Use Difference View}
%\com{d) Clarify using confounder (Var ~ x)}
%\com{3. Find interesting pane.}
%\com{4. Get Details on Demand.}
%%
The analysis workflow follows the Visual Analytics Mantra of Keim et al \cite{Keim2008}:

\textbf{Analyze First.} Choosing an initial regression formula triggers the \emph{Regression Cube} calculation on the given data set, filtering the dimensions of the dependent feature through the CFS algorithm.

\textbf{Show the Important.} The 3D-cube visualization acts as overview over the whole data set.
%%
Regression models with large R$^2$ values can be spotted fast here, steering the attention to the respective slice.

\textbf{Zoom, Filter and Analyze Further.} The slices of interest can then be analyzed using the 2D mosaic plot of the slice.

\textbf{Details on Demand.} Precise information about the individual regression models (coefficients, associated confidence intervals and p-values) can be retrieved based on the data point representatives (e.g. in a hover-modal on a currently selected data point).
\\\\
As seen in Fig~\ref{fig:Workflow}, the workflow is highly iterative.
%%
Observations in the 2D-mosaic plot or simply the CFS-based features can trigger new analyses by adjusting the underlying regression formulas.
%%
This can be carried out either to refine the current formula based on observations, or creating a new regression cube for a difference view.
%%
\paragraph{Formula Types.}
%%
Input formulas reflect \emph{hypotheses} about the data.
%%
Using the operators, dynamic variables and dataset features, many different assumptions can be formulated.
%%
Here we show just a few examples on how formulas can serve different purposes.

$Z \sim X + Y$ is the default if the user did not specify a formula.
%%
This cube represents all possible combinations of two independent features towards all features in the data set, since we do not know which feature(s) are of interest.
%%
Each slice represents a different target feature.
%%
It is therefore suitable for a \emph{hypothesis-free} explorative analysis to give a general impression about relationships in the data set.

$Cancer \sim X + Y + Z + feature_1:feature_2$ is an example for the formulation of a hypothesis, where a specific feature is analyzed.
%%
The interaction between $feature_1$ and $feature_2$ is also included in all regression formulas.
%%

$Cancer \sim X + Y + Z$ subtracted with results from $Cancer \sim Age$ excludes the confounding effect age has towards the target $Cancer$ feature.
%%
\com{TODO: divide using analysis approaches, e.g. Hypothesis-based \& Hypothesis-free with follow up on the formulas used?}
%%
\paragraph{Cube Comparison.}

\section{System Design and Implementation} \label{Interaction- and Visualization Techniques}
%%
